---
title: "Oregon LiDAR CHM Mapping Project"
author: "Gavin Adams - Adams Geospatial"
output:
  html_document:
    df_print: paged
---
```{r}
#devtools::install_github("sailbot24/oRLiDAR")
library(oRLiDAR)
library(terra)
library(sf)
library(dplyr)
```

```{r}
year_filter <- 1946
size_filter <- 9 

filtered_veg <- st_read("Veg_ShapeFiles/2003-Veg_Siuslaw/veg.shp", quiet = T) %>%
  mutate(YR_ORIG = as.numeric(YR_ORIG),L1_SZCL = as.numeric(L1_SZCL)) %>% 
  filter(YR_ORIG > year_filter & L1_SZCL >size_filter)

for(i in 1:nrow(filtered_veg)){
  fro
}


```


# Packages

Like any R project, the first step is to load in the packages that are required for this project. The one thing of note is the last line loads in the CHM downloading function. As described in the comments in the cell, the path to the CHM script must be included for R to pull the functions in. 

```{r}

library(dplyr)

library(terra)
library(sf)

library(ForestTools)

library(whitebox) # This package is worth looking in to. 

#source("oregon_chm_downloader.r") # The file "oregon_chm_downlodader.r" needs to be in the same folder as the .RMD otherwise a path to the file needs to be specified
#TODO turn the CHM downloaded in to a R package
```

# Filter the Veg Shapefiles

This next block of code sets the filters for the shape file. Then reads in the vegetation shape file. (you may need to update paths to the shapefile if you try to run the script)

```{r}
year_filter <- 1946
size_filter <- 9 

veg_shapefiles <- st_read("Veg_ShapeFiles/2003-Veg_Siuslaw/veg.shp", quiet = T) 
filtered_veg <- veg_shapefiles %>% filter(YR_ORIG > year_filter & L1_SZCL >size_filter)

```

# Downloading the First AOI CHM Raster
Here is the test of the first AOI based on the filtered vegetation shapefiles. The AOI is then fed in to the downloader function. When the mask parameter is set to true the output raster will be masked to the shapefile. There is also a parameter that allows for the definition of the output path of the raster. I am going to go in and add an parameter to not save a copy of the raster locally when its downloaded. 

```{r}
source("oregon_chm_downloader.r")
aoi <- filtered_veg[1,]
chm<- download_oregon_chm(aoi, mask = TRUE) #TODO add an option to not save the downloaded raster

```

# Running  Forest Tools on the First AOI

Here is the running of forest tools. The lin function would need to be adjusted to better match the correct function for the AOI. Once we find a function that we like for the VWF we shoudl have an accurate tree count without over sampeling. 
```{r}

minHeight = 12
# Function for defining dynamic window size
lin <- function(x){x * 0.1 + 1}

# Detect treetops
ttops <- vwf(chm, winFun = lin, minHeight = minHeight)
# Plot CHM
plot(chm, xlab = "", ylab = "", xaxt='n', yaxt = 'n')
# Add dominant treetops to the plot
plot(ttops$geometry, col = "blue", pch = 20, cex = 0.5, add = TRUE)

# Create polygon crown map
crowns_poly <- mcws(treetops = ttops, CHM = chm, format = "polygons", minHeight =minHeight)
st_write(ttops, "aoi1_tt.gpkg", delete_dsn = TRUE, quiet = T)
```

## mapping the results of forest tools

```{r}
#  Map CHM
library(tmap)

# Get exact range and add buffer to highest value
height_range <- range(values(chm), na.rm = TRUE)
max_height <- ceiling(height_range[2] + 10)  # Add buffer to include all values
breaks <- seq(0, max_height, by = 10)

tm_shape(chm) +
  tm_raster(
    col.scale = tm_scale(values = "viridis",
                        breaks = breaks,
                        outliers.rm = TRUE),
    col.legend = tm_legend(title = "Canopy Height (ft)")) +
  tm_shape(crowns_poly) + 
  tm_borders(col = "white", 
             lwd = 0.5, 
             fill_alpha = 0.5) +
  #tm_graticules() +
  tm_compass(
    type = "arrow", 
    position = c("left", "bottom"),
    size = 1.2
  ) +
  tm_scale(
    position = c("left", "bottom"),
    offset.x = 1.5
  ) +
  tm_title("Canopy Height Model",
           position = c("center", "top")) +
  tm_layout(
    frame = FALSE,
    #outer.margins = c(0.15, 0.15, 0.15, 0.15),  # Increased margins
    inner.margins = c(0.1, 0.1, 0.1, 0.1)
  )
```

# Finding the Optimal Search Window

```{r, include = FALSE}
# Function to test different window parameters
test_window_functions <- function(chm, plot_results = TRUE) {
    # Define different window functions
    win_functions <- list(
        conservative = function(x){x * 0.08 + 1},  # Smaller windows
        moderate = function(x){x * 0.12 + 1.5},    # Medium windows
        aggressive = function(x){x * 0.15 + 2}     # Larger windows
    )
    
    # Test each function
    results <- list()
    for(name in names(win_functions)) {
        ttops <- vwf(chm, winFun = win_functions[[name]], minHeight = minHeight)
        results[[name]] <- ttops
        
        if(plot_results) {
            # Plot results
            plot(chm, main = paste("Window Function:", name))
            plot(ttops$geometry, col = "blue", pch = 20, cex = 0.5, add = TRUE)
        }
    }
    
    # Print summary statistics
    for(name in names(results)) {
        cat(sprintf("\n%s function found %d trees", name, nrow(results[[name]])))
    }
    
    return(results)
}

# Run the test
test_results <- test_window_functions(chm)

# Calculate average crown diameter for validation
validate_crowns <- function(ttops, chm) {
    crowns <- mcws(treetops = ttops, CHM = chm, format = "polygons", minHeight = minHeight)
    crown_areas <- st_area(crowns)
    avg_diameter <- mean(sqrt(crown_areas/pi) * 2)
    return(list(
        n_trees = nrow(ttops),
        avg_diameter = avg_diameter,
        crowns = crowns
    ))
}

# Optimize based on typical crown diameter-to-height relationships
optimize_window <- function(chm) {
    # Get height distribution
    heights <- values(chm)
    heights <- heights[!is.na(heights)]
    
    # Calculate expected crown widths based on height
    # Using general relationship: crown width ≈ 0.15-0.25 * height for conifers
    # Adjust these coefficients based on your forest type
    mean_height <- mean(heights, na.rm = TRUE)
    expected_crown_width <- mean_height * 0.2  # Using 0.2 as middle ground
    
    # Convert to window function parameters
    # Window should be about half the crown width at mean height
    slope <- (expected_crown_width/2) / mean_height
    intercept <- expected_crown_width/4  # Base window size
    
    # Return optimized function
    return(function(x) x * slope + intercept)
}

# Get optimized window function
opt_win <- optimize_window(chm)

# Test optimized function
ttops_opt <- vwf(chm, winFun = opt_win, minHeight = minHeight)
crowns_opt <- mcws(treetops = ttops_opt, CHM = chm, format = "polygons", minHeight = minHeight)

# Validate results
plot(chm, main = "Optimized Window Function")
plot(ttops_opt$geometry, col = "blue", pch = 20, cex = 0.5, add = TRUE)
plot(crowns_opt, border = "white", add = TRUE)

# Print statistics
cat(sprintf("\nOptimized function found %d trees", nrow(ttops_opt)))
cat(sprintf("\nAverage crown diameter: %.2f feet", 
            mean(sqrt(st_area(crowns_opt)/pi) * 2)))
```

```{r, include= FALSE}
# Function to compare different window functions
compare_window_functions <- function(chm, minHeight) {
    # Define different window functions from literature
    win_functions <- list(
        kane = function(x){x * 0.15 + 2},     # Kane et al. 2010
        popescu = function(x){x * 0.13 + 1.5}, # Popescu & Wynne 2004
        kini = function(x){x * 0.11 + 1.2},    # Kini & Popescu 2004
        chen_df = function(x){x * 0.18 + 1.8}, # Chen et al. 2006 (Douglas-fir)
        chen_pp = function(x){x * 0.225 + 2}   # Chen et al. 2006 (Ponderosa Pine)
    )
    
    # Store results
    results <- list()
    
    # Create plot layout
    par(mfrow = c(2, 3))
    
    # Test each function
    for(name in names(win_functions)) {
        # Detect trees
        ttops <- vwf(chm, winFun = win_functions[[name]], minHeight = minHeight)
        
        # Create crowns
        crowns <- mcws(treetops = ttops, CHM = chm, format = "polygons", minHeight = minHeight)
        
        # Store results
        results[[name]] <- list(
            ttops = ttops,
            crowns = crowns,
            n_trees = nrow(ttops),
            mean_height = mean(ttops$height),
            mean_crown_area = mean(st_area(crowns))
        )
        
        # Plot results
        plot(chm, main = paste("Method:", name, "\nTrees:", nrow(ttops)))
        plot(crowns, border = "white", add = TRUE)
        plot(ttops$geometry, col = "blue", pch = 20, cex = 0.5, add = TRUE)
    }
    
    # Reset plot layout
    par(mfrow = c(1, 1))
    
    # Print summary statistics
    cat("\nSummary Statistics:\n")
    for(name in names(results)) {
        cat(sprintf("\n%s method:", name))
        cat(sprintf("\n  Trees detected: %d", results[[name]]$n_trees))
        cat(sprintf("\n  Mean height: %.2f ft", results[[name]]$mean_height))
        cat(sprintf("\n  Mean crown area: %.2f sq ft\n", results[[name]]$mean_crown_area))
    }
    
    return(results)
}

# Create visualization of window functions
plot_window_functions <- function() {
    heights <- seq(0, 100, by = 10)
    
    plot(heights, heights * 0.15 + 2, type = "l", col = "red",
         xlab = "Tree Height (ft)", ylab = "Window Radius (ft)",
         main = "Window Functions Comparison")
    lines(heights, heights * 0.13 + 1.5, col = "blue")
    lines(heights, heights * 0.11 + 1.2, col = "green")
    lines(heights, heights * 0.18 + 1.8, col = "purple")
    lines(heights, heights * 0.225 + 2, col = "orange")
    legend("topleft", 
           legend = c("Kane et al.", "Popescu & Wynne", "Kini & Popescu", 
                     "Chen et al. (DF)", "Chen et al. (PP)"),
           col = c("red", "blue", "green", "purple", "orange"),
           lty = 1)
}

# Print citations
print_citations <- function() {
    cat("
Citations:

1. Kane, V. R., McGaughey, R. J., Bakker, J. D., Gersonde, R. F., Lutz, J. A., & Franklin, J. F. (2010). 
   Comparisons between field-and LiDAR-based measures of stand structural complexity. 
   Canadian Journal of Forest Research, 40(4), 761-773.

2. Popescu, S. C., & Wynne, R. H. (2004). 
   Seeing the trees in the forest: Using lidar and multispectral data fusion with local filtering and 
   variable window size for estimating tree height. 
   Photogrammetric Engineering & Remote Sensing, 70(5), 589-604.

3. Kini, A. U., & Popescu, S. C. (2004). 
   TreeVaW: A versatile tool for analyzing forest canopy LIDAR data: A preview with an eye towards future. 
   In Proceedings of the ASPRS 2004 Fall Conference.

4. Chen, Q., Baldocchi, D., Gong, P., & Kelly, M. (2006). 
   Isolating individual trees in a savanna woodland using small footprint lidar data. 
   Photogrammetric Engineering & Remote Sensing, 72(8), 923-932.
    ")
}

# Create detailed map for best method
create_detailed_map <- function(results, method_name, chm) {
    # Get results for specified method
    method_results <- results[[method_name]]
    
    # Get height range for legend
    height_range <- range(values(chm), na.rm = TRUE)
    breaks <- seq(0, ceiling(height_range[2]), by = 10)
    
    # Create map
    tm_shape(chm) +
        tm_raster(
            col.scale = tm_scale(values = "viridis",
                               breaks = breaks),
            col.legend = tm_legend(title = "Canopy Height (ft)")) +
        tm_shape(method_results$crowns) + 
        tm_borders(col = "white", 
                  lwd = 0.5, 
                  fill_alpha = 0.5) +
        tm_scalebar(position = c("left", "bottom"),
                   width = 0.15) +
        tm_compass(type = "arrow", 
                  position = c("left", "bottom"),
                  size = 1,
                  offset.x = 0.2) +
        tm_title(paste(method_name, "Method Results"),
                position = c("center", "top")) +
        tm_layout(frame = TRUE,
                 outer.margins = c(0.15, 0.15, 0.15, 0.15),
                 inner.margins = c(0.1, 0.1, 0.1, 0.1))
}

# Run the analysis
# Assuming you have your CHM loaded as 'chm' and minHeight defined
# minHeight <- 6  # Example minimum height threshold

# Run the comparison
results <- compare_window_functions(chm, minHeight)

# Plot window functions
plot_window_functions()

# Print citations
print_citations()

# Create detailed map for best method (example using Kane method)
detailed_map <- create_detailed_map(results, "kane", chm)
```

```{r}
library(ggplot2)

# Convert to factor if needed
veg_shapefiles$`L1_SPP1` <- as.factor(veg_shapefiles$`L1_SPP1`)

# Define custom labels
species_labels <- c("ABPR" = "ABPR\nnoble fir", "ACMA" = "ACMA\nbigleaf maple", 
                    "ALRU" = "ALRU\nred alder", "PICO" = "PICO\nshore pine", 
                    "PISI" = "PISI\nSitka spruce", "PSME" = "PSME\nDouglas-fir", 
                    "THPL" = "THPL\nWestern red-cedar", "TSHE" = "TSHE\nWestern hemlock", 
                    "TREE" = "TREE\nunknown tree species", "TREED" = "TREED\nUnknown deciduous tree species", 
                    "TREEC" = "TREEC\nUnknown conifer tree species")

ggplot(filtered_veg, aes(x = `L1_SPP1`)) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(labels = species_labels) +  # Apply custom labels
  labs(title = "The Most Common Species in the Layer in the Filtered Data Set", x = "Species", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate labels for readability

```

# Running for All AOIs 

```{r}

```

# Source the tree age calculator
source("oregon_tree_age_calculator.R")

# Load National Forest boundaries
nf_path <- "Veg_ShapeFiles/FS_National_Forests_Dataset_(US_Forest_Service_Proclaimed_Forests)/FS_National_Forests_Dataset_(US_Forest_Service_Proclaimed_Forests).shp"
national_forests <- st_read(nf_path, quiet = TRUE)

# Filter for Siuslaw National Forest
siuslaw <- national_forests %>% 
  filter(FORESTNAME == "Siuslaw National Forest")

# Create hexagon grid (50 acres each)
hex_grid <- create_hexagon_grid(siuslaw, area_acres = 50)

# Process just the first hexagon
sample_hexagon <- hex_grid[1, ]
sample_result <- process_hexagon(sample_hexagon, 1)

# View the results
print(sample_result)

